Experiences with Planning for Natural Language Generation

### 2008-07-25 trunk/Papers/experiences/conclusion.tex
small changes

In this paper, we introduced two novel planning domains arising from
problems in natural language generation: the sentence generation
domain and the GIVE navigation domain. We also reported on the results
of a number of experiments in which we applied off-the-shelf planners
to these domains.

Our results were mixed. While modern planners do a pretty good job of
controlling the complexity of search, they also suffer from practical
problems that limit their performance in unexpected ways. In
particular, in both domains the grounding step performed by both
planners dominates the time it takes to perform the search itself. In
future work, it would be interesting to explore planners such as SHOP2
\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the
initial grounding step.  We are also considering to propose some of
our domains as IPC benchmarks.

The NLG community's recent interest in planning presents a valuable
opportunity for planning researchers: provided some of the challenges
we have highlighted can be addressed, projects like GIVE (and other
NLG-inspired problem domains) offer a constructive platform for the
planning community to showcase their techniques to a wider
audience---and to improve the quality of their tools for real-world
planning tasks.  Furthermore, GIVE highlights planning-related
challenges such as plan execution monitoring and plan presentation
and summarisation and offers a platform on which such technologies can
be evaluated in experiments with human users.

>>>>

In this paper, we introduced two novel planning domains arising from
problems in natural language generation: the sentence generation
domain and the GIVE navigation domain. We also reported on the results
of a number of experiments in which we applied off-the-shelf planners
to a set of problem instances in these domains.

Our results were mixed. While modern planners do a pretty good job of
controlling the complexity of search, they also suffer from practical
problems that limit their performance in unexpected ways. In particular, in
both domains the grounding step performed by FF and SGPLAN dominates the
time it takes to perform the search itself. In future work, we hope to
extend our analysis by exploring other planners, such as SHOP2
\cite{DBLP:journals/jair/NauAIKMWY03}, which do not perform the initial
grounding step. We also believe our domains would provide suitable
challenges for planners entered in future editions of the IPC.
%We are also considering to propose some of our domains as IPC benchmarks.

The NLG community's recent interest in planning presents a valuable
opportunity for planning researchers.  While GIVE highlights
planning-related challenges such as plan execution monitoring and plan
presentation (i.e., summarisation and elaboration), it also offers a
platform on which such technologies can be evaluated in experiments with
human users.  Furthermore, more mainstream research on reasoning under
uncertainty, planning with incomplete information and sensing, and resource
management can also be applied in this domain setting. Provided some of the
challenges we have highlighted can be addressed, projects like GIVE (and
other NLG-inspired problem domains) offer a constructive platform for the
planning community to showcase their techniques to a wider audience---and
to improve the quality of their tools for real-world planning tasks.

%The NLG community's recent interest in planning presents a valuable
%opportunity for planning researchers: provided some of the challenges
%we have highlighted can be addressed, projects like GIVE (and other
%NLG-inspired problem domains) offer a constructive platform for the
%planning community to showcase their techniques to a wider
%audience---and to improve the quality of their tools for real-world
%planning tasks.  Furthermore, GIVE highlights planning-related
%challenges such as plan execution monitoring and plan presentation
%and summarisation and offers a platform on which such technologies can
%be evaluated in experiments with human users.

### 2008-07-25 trunk/Papers/experiences/discussion.tex
small changes

The positive conclusion we can draw from the simple experiments
reported above is that modern planners are fairly good at controlling
the search for many of the moderately-sized NLG problems we looked
at. This is particularly true for FF in the sentence generation
domain, which generates nontrivial 14-word sentences in about two
seconds. There is still room for improvement, however; special-purpose
algorithms generate much larger referring expressions in milliseconds
\cite{AreKolStr08}.  Although search efficiency is not the main
concern in these domains, it nevertheless becomes a factor in
scenarios like Experiment~2, where the world is almost completely
unconstrained and (unsurprisingly) the search becomes much slower.

The most restrictive bottleneck in the two domains we investigated is
the initial grounding step that many modern planners perform. While
this may be a good strategy for traditional planning benchmarks and
IPC domains---where plan length often dominates universe size, and an
initial investment into grounding pays off in saved instantiations
during the search---this approach is less effective in the domains
we've considered. As our experiments have shown, there are natural
planning domains in which relatively short plans must be computed in
large universes. For instance, in the case of the sentence generation
domain, it is not unrealistic to look for plans of length 20 over a
universe consisting of thousands of domain individuals and tens of
thousands of actions, some of which require three domain individuals
as parameters. The common strategy of splitting the universe over
several types of individuals will not be very effective here.

The recent trend in planning research has (rightfully) focused on the
development of algorithms that control search in sophisticated ways,
resulting in a host of planners that are more powerful and more
successful than their predecessors. However, the common strategy of
grounding out sets of predicates and actions in implementations of
these algorithms has a very pronounced effect on domains such as those
described above. Reachability analysis has the potential to help---we
could gain a lot by only computing reachable instances---, but this
too is typically done on the grounded problem. \todo{check if this is
  true} Since our domains are not that unusual in their structure and
composition, we hope that the lessons learnt from our experiences can
help improve the performance of current systems. We believe that such
improvements are necessary if planning is to become a more mature
technology that can offer tools to a wider community of users. At the
end of the day, real-world users will care about the \emph{total}
runtime of a planner, and this is more than just the search time.

>>>>

The positive conclusion we can draw from the simple experiments
reported above is that modern planners are fairly good at controlling
the search for many of the moderately-sized NLG problems we looked
at. This is particularly true for FF in the sentence generation
domain, which generates nontrivial 14-word sentences in about two seconds.
There is still room for improvement, however: special-purpose
algorithms generate much larger referring expressions in
milliseconds \cite{AreKolStr08}.  Although search efficiency is not the
main concern in these domains, it nevertheless becomes a factor in
scenarios like Experiment~2, where the world is almost completely
unconstrained and (unsurprisingly) the search becomes much slower.

The most restrictive bottleneck in the two domains we investigated is
the initial grounding step that many modern planners perform. While
this may be a good strategy for traditional planning benchmarks and
IPC domains---where plan length often dominates universe size, and an
initial investment into grounding pays off in saved instantiations
during the search---this approach is less effective in the domains
we've considered. As our experiments show, there are natural
planning domains in which relatively short plans must be computed in
large universes. For instance, in the case of the sentence generation
domain, it is not unrealistic to look for plans of length 20 over a
universe consisting of thousands of domain individuals and tens of
thousands of actions, some of which require three domain individuals
as parameters. The common strategy of splitting the universe over
several types of individuals will not be very effective here.

The recent trend in planning research has (rightfully) focused on the
development of algorithms that control search in sophisticated ways,
resulting in a host of planners that are more powerful and more
successful than their predecessors. However, the common strategy of
grounding out sets of predicates and actions in implementations of these
algorithms has a very pronounced effect on domains such as those described
above. More sophisticated techniques based on reachability analysis may
have the potential to help improve this situation, provided such
approaches can avoid the drawbacks resulting from unnecessary grounding.
Since our domains are not that unusual in their structure and composition,
we hope that the lessons learnt from our experiences can help improve the
performance of current systems.  We believe that such improvements are
necessary if planning is to become a more mature technology that can offer
tools to a wider community of users.  At the end of the day, real-world
users will care about the \emph{total} runtime of a planner, and this is
more than just the search time.
%has the potential to help---we could gain a lot by only computing
%reachable instances---, but this too is typically done on the grounded
%problem.  \todo{check if this is true}

### 2008-07-25 trunk/Papers/experiences/domains.tex
small changes

Once the appropriate actions are defined, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.

@@

A map of an example GIVE world is shown in
Figure~\ref{fig:give-development-world}.  In this world, the user's
task is to pick up a trophy in the top left room.  The trophy is
hidden in a safe behind a picture, so the user must first move the
picture out of the way and open the safe by pushing a sequence of
buttons (the small square boxes on the walls) in the correct order.
In order to get to all these buttons, the user must also open the door
in the centre of the map and deactivate the alarm tile by pushing
further buttons.  To simplify both the planning and the NLG task, the
world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one
tile to the centre of the next. Figure~\ref{fig:give-planning} shows
some of the available actions in PDDL syntax.

@@

The plans generated in the GIVE domain are often nontrivial. For instance,
in the example above the shortest plan consists of 108 action steps; the
first few steps are as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north}, \mathsf{west})$
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$
\end{enumerate}

@@

The GIVE task comes with a number of interesting research challenges
both for the NLG and the planning communities.  On the NLG side, one
major problem is that the natural language instruction sequences can't
simply be obtained by verbalising each action instance in a plan in
turn.  For instance, the NLG system could generate an instruction
sequence starting with ``turn left; walk forward; press the button;
turn right; walk forward; walk forward; turn right; walk forward; walk
forward; turn left; walk forward''.  However, this would be clumsy and
boring, and it is to be expected that frustrated users will quickly
cancel an interaction with such a system, resulting in a bad
evaluation score.  Having just pressed the button, it would be much
better to say ``turn around and walk through the door''. That is,
multiple action instances should be merged into a single instruction;
plans must be \emph{summarised} in order to present them to the
user. Conversely, it may also be necessary to express a single
planning step with several instructions.  Having just entered the top
left room, it may be easier for the user to understand the instruction
``walk to the centre of the room; turn right; now press the green
button in front of you'' than the instruction ``press the green button
on the wall to your right''.  To plan the referring expression ``the
green button in front of you'' at a time when the button is not yet in
front of the user, the NLG system must keep track of the hypothetical
changes in what objects are visible to the user.  Thus the acts of
referring and instructing, and the modules for planning and NLG, must
be tightly integrated.

>>>>

Once the appropriate actions are defined, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$,
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$,
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$.
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the
generated sentence ``the white rabbit sleeps'', can be systematically
reconstructed. Thus, we can solve the sentence generation problem via
the detour through planning and bring current search heuristics for
planning to bear on generation.

@@

A map of an example GIVE world is shown in
Figure~\ref{fig:give-development-world}.  In this world, the user's
task is to pick up a trophy in the top left room.  The trophy is
hidden in a safe behind a picture, so the user must first move the
picture out of the way and open the safe by pushing a sequence of
buttons (the small square boxes on the walls) in the correct order.
In order to get to all these buttons, the user must also open the door
in the centre of the map and deactivate the alarm tile by pushing
further buttons.  To simplify both the planning and the NLG task, the
world is discretised into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one
tile to the centre of the next. Figure~\ref{fig:give-planning} shows
some of the available actions in PDDL syntax.

@@

The plans generated in the GIVE domain are often nontrivial. For instance,
in the above example the shortest plan consists of 108 action steps; the
first few steps are as follows:
%
\begin{enumerate}
\item $\mathsf{turn}\textsf{-}\mathsf{left}(\mathsf{north},
\mathsf{west})$,
\item $\mathsf{move}(\mathsf{pos\_5\_2}, \mathsf{pos\_4\_2}, \mathsf{west})$,
\item $\mathsf{manipulate}\textsf{-}\mathsf{b1}\textsf{-}\mathsf{off}\textsf{-}\mathsf{on}(\mathsf{pos\_5\_2})$,
\item $\mathsf{turn}\textsf{-}\mathsf{right}(\mathsf{west}, \mathsf{north})$.
\end{enumerate}

@@

The GIVE task comes with a number of interesting research challenges
both for the NLG and the planning communities.  On the NLG side, one
major problem is that the natural language instruction sequences can't
simply be obtained by verbalising each action instance in a plan in turn.
For instance, the NLG system could generate an instruction sequence
starting with ``turn left; walk forward; press the button; turn right; walk
forward; walk forward; turn right; walk forward; walk forward; turn left;
walk forward''.  However, such directions would be clumsy and boring, and
it is to be expected that frustrated users will quickly cancel an
interaction with such a system, resulting in a bad evaluation score.
Having just pressed the button, it would be much better to say ``turn
around and walk through the door''. That is, plans must be
\emph{summarised} by merging multiple action instances into a single
instruction, in order to present them to the user. Conversely, it may also
be necessary to \emph{elaborate} on a single planning step by expressing it
with several instructions.  Having just entered the top left room, it may
be easier for the user to understand the instruction ``walk to the centre
of the room; turn right; now press the green button in front of you'',
rather than the instruction ``press the green button on the wall to your
right''.  To plan the referring expression ``the green button in front of
you'' at a time when the button is not yet in front of the user, the NLG
system must keep track of the hypothetical changes in what objects are
visible to the user.  Thus the acts of referring and instructing, and the
modules for planning and NLG, must be tightly integrated.

### 2008-07-25 trunk/Papers/experiences/experiences.tex
added abstract
spellcheck
small changes

\begin{abstract}
We investigate the application of modern planning techniques to domains
arising from problems in natural language generation (NLG). In particular,
we consider two novel NLG-inspired planning problems, the sentence
generation domain and the GIVE (``Generating Instructions in Virtual
Environment'') domain, and investigate the efficiency of FF and SGPLAN in
these domains. We also compare our results against an ad-hoc implementation
of GraphPlan in Java. Our results are mixed. While modern planners are able
to quickly solve many moderately-sized instances of our problems, the
overall planning time is dominated by the grounding step that these
planners perform (rather than search). This has a pronounced effect on our
domains which require relatively short plans in large universes.  We share
our experiences and offer these domains as challenges for the planning
community.
\end{abstract}

>>>>

\begin{abstract}
We investigate the application of modern planning techniques to domains
arising from problems in natural language generation (NLG). In particular,
we consider two novel NLG-inspired planning problems, the sentence
generation domain and the GIVE (``Generating Instructions in Virtual
Environment'') domain, and investigate the efficiency of FF and SGPLAN in
these domains. We also compare our results against an ad-hoc implementation
of GraphPlan in Java. Our results are mixed. While modern planners are able
to quickly solve many moderately-sized instances of our problems, the
overall planning time is dominated by the grounding step that these
planners perform, rather than search. This has a pronounced effect on our
domains which require relatively short plans but have large universes. We
share our experiences and offer these domains as challenges for the
planning community.
\end{abstract}

### 2008-07-25 trunk/Papers/experiences/introduction.tex
small changes

We then discuss some of our experiences using off-the-shelf planners
in our two domains. In particular, we explore the efficiency of FF
\cite{HoffmannNebel01} and SGPLAN
\cite{hsu06:_new_featur_in_sgplan_for}---planners that are easily
available, support expressive preconditions and effects out of the
box, and have been successful on traditional benchmarks, and in the
International Planning Competition (IPC)---on a range of problem
instances from our planning domains.  We also compare these results
against an ad-hoc Java implementation of GraphPlan \cite{Blum1997}.
Our findings are mixed. On the one hand, it turns out that modern
planners handle some of the search problems that arise in NLG quite
easily (although large problem instances still remain a challenge).
On the other hand, these same planners spend tremendous amounts of
time on preprocessing. For instance, FF spends 90\% of its runtime in
the sentence generation domain on grounding out literals and actions,
and ends up slower than our version of GraphPlan that avoids
grounding. For domains like ours, which are dominated by the number of
actions and the universe size, rather than the combinatorics of the
search problem, this observation suggests that the overall runtime of
a planner could be improved by grounding more selectively. We
therefore offer these domains as challenges for the planning
community.

>>>>

We then discuss some of our experiences using off-the-shelf planners in our
two domains. In particular, we explore the efficiency of FF
\cite{HoffmannNebel01} and SGPLAN
\cite{hsu06:_new_featur_in_sgplan_for}---planners that are easily
available, support an expressive subset of PDDL for encoding preconditions
and effects, and have been successful on traditional benchmarks and in the
International Planning Competition (IPC)---on a range of problem instances
from our planning domains.  We also compare these results against an ad-hoc
Java implementation of GraphPlan \cite{Blum1997}.  Our findings are mixed.
On the one hand, it turns out that modern planners handle some of the
search problems that arise in NLG quite easily (although large problem
instances still remain a challenge).  On the other hand, these same
planners spend tremendous amounts of time on preprocessing. For instance,
FF spends 90\% of its runtime in the sentence generation domain on
grounding out literals and actions, and ends up slower than our version of
GraphPlan that avoids grounding. For domains like ours, which are dominated
by the number of actions and the universe size, rather than the
combinatorics of the search problem, this observation suggests that the
overall runtime of a planner could be improved by grounding more
selectively. We therefore offer these domains as challenges for the
planning community.
%available, support expressive preconditions and effects out of the box,
